NumPy---Array Computation for Python

### 2019-12-20 body.tex
Proofread manuscript

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Since its first stable release (1.0) in October 2006 it has unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
Not surprisingly, many important scientific research projects and
results rely on NumPy.
It plays an essential role in many research analysis pipelines in
fields as diverse as physics, biology, astronomy, neuroscience,
material science, engineering, and chemistry.
% w/ citations
In particular, NumPy played a fundamental role in discovering
gravitational waves\cite{pycbc}, imaging black
holes\cite{eht-imaging}, and exploring the solar system (... Kepler? https://ui.adsabs.harvard.edu/abs/2012ascl.soft08004S/abstract).

...
%[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
%  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
%  IDL is now below Python, Pangeo <-> xarray, Earthcube,
%  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
%  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
%  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]

@@

We showed that NumPy is actively developed according to a community
vetted plan that integrates formal enhancement proposals.  Next, we
give a more in-depth overview of the team developing NumPy, their
activities, as well as the governance and community structures that
guide the collaboration.

@@

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these 17 maintainers were active in 2019.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through github
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
With each release attracting more than a hundred new contributors.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
    \caption{Number of pull requests merged into the NumPy master branch for each
        quarter since 2012. The total number of PR is indicated with the
        lower blue area showing the portion contributed by current or previous
        maintainers.}\label{fig:prs-over-time}
\end{figure}

The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
PR merged into the NumPy master branch.
In this plot PR opened by current and old maintainers are indicated.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.
Over the course of its history, NumPy has attracted PR by 823 contributors.
However, its development still relies heavily on a relatively small number
of active maintainers, who share more than half of the contributions among
themselves.

% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
% https://github.com/numpy/numpy/pull/6352
NumPy adopted an official Governance Document on October~5,
2015\cite{NumPyProjectGovernance}.
Project decisions are usually made by consensus of interested contributors.
This means for many thing everyone is entrusted with veto power.
A Steering Council, currently composed of 12~members, facilitates this
process and oversees daily development of the project by contributing code
and reviewing contributions from the community.

@@

The NumPy array data structure and universal functions, which we describe
below, make up the core of NumPy.
The NumPy library also includes
a set of library functions for manipulating arrays and doing scientific computation,
libraries for unit tests and Python package building,
and the program \code{f2py}\cite{peterson2009f2py} for wrapping Fortran code in Python.

@@

The \code{ndarray} data structure stores regularly strided homogeneous data
types in a contiguous block memory, allowing for the efficient representation
of $n$-dimensional data.
A \emph{masked array} class, built on top of the \code{ndarray}, is
provided as well. This class allows elements of an array to be masked,
indicating that the data is missing or unknown.
More details about the data structure are given in ``The NumPy array:
a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.

The \emph{universal functions} (or \emph{ufuncs})
are functions written in C that implement efficient looping over
NumPy arrays. An important feature of ufuncs is the built-in
implementation of \emph{broadcasting}.  For example, the function
\code{arctan2(x, y)} is a ufunc that accepts two values and computes
$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
the ufunc will take care of looping over the dimensions of the inputs
in such a way that if, say, \code{x} is a 1-d array with length 3, and
\code{y} is a 2-d array with shape $2 \times 1$, the output will be
an array with shape $2 \times 3$.  The ufunc machinery takes care
of calling the function with all the appropriate combinations of
input array elements to complete the output array.
The elementary arithmetic operations of addition, multiplication, etc.,
are implemented as ufuncs, so broadcasting applies even in expressions
such as \code{x + y * z}.

@@

NumPy provides a large library of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
For historical reasons, NumPy also includes basic functionality for
linear algebra,
fast Fourier transforms and windowing,
and polynomial fitting.
We will keep supporting these, but we will not expand beyond them.

@@

Being at the base of the scientific Python stack, NumPy provides some basic
infrastructure for other packages in the scientific Python ecosystem.
The \code{numpy.testing} subpackage provides functions such as
\code{assert\_allclose(actual, desired)} that may be used in unit
test suites for code that uses NumPy arrays.
The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
BLAS/LAPACK, and other relevant libraries for scientific computing.
The program \code{f2py} is a tool for
building NumPy-aware Python wrappers of Fortran functions.
NumPy itself does not use any Fortran code;  F2PY is part of NumPy
for historical reasons.

@@

NumPy has, throughout its lifetime, seen continuous development.
E.g., bugs fixes, minor improvements, cleanups, and releases have
happened regularly.  However, with the NEP process in place we now
have the flexibility to plan and implement changes of larger scope.
We highlight two of those below, as well as changes made to our
testing infrastructure to support hardware platforms used in large
scale computing.

@@

The central position of NumPy takes within the numerical and scientific
Python ecosystem means that a vast number of projects are built on it.
These projects are consumers of the NumPy API.

On the other side, projects are providers of a \emph{NumPy-like API} and array
objects targeting audiences with specialized needs beyond NumPy's capabilities.
For example, the NumPy API is implemented by
several popular tensor computation libraries including
CuPy\footnote{\url{https://cupy.chainer.org/}},
Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
and to a lesser extent
TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
It is also implemented in packages that support sparse arrays
such as \code{scipy.sparse} and \code{pydata.sparse}.
Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.

The multitude of specialized projects creates the difficulty that consumers
of these NumPy-like APIs write code specific to a single project and do not support
all of the above array providers.
This is a burden for users relying on the specialized Array-like, since
a tool they need may not work for them.
Additionally, a user who has the need of transitioning from NumPy to a more
specialized array, is faced with a challenge.
These issue can lead into a fragmentation of the user bases into consumers
of different NumPy-like APIs.

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of Array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.  For
example, here is how a CuPy GPU array can be passed through NumPy for
processing, with all operations being dispatched back to CuPy:

@@

\begin{lstlisting}
from numpy.random import Generator, PCG64, SeedSequence
seq = SeedSequence(10304245474441179933310169597522947680)
pcg = PCG64(seq)
gen = Generator(pcg)
\end{lstlisting}

@@

The final goal of the new API is to improve extensibility. \code{RandomState} is
a monolithic object that obscures all of the underlying state and functions. The
component architecture is one part of the extensibility improvements. The
underlying functions (written in C) which transform the output of a bit
generator to other distributions are available for use in CFFI. This allows the
same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
numba. Both the bit generators and the low-level functions can also be used in C
or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
NumPy source. Alternative approaches that avoid this extra step are being
explored.}

@@

Most of the technical work for the Moore and Sloan grant is now complete.
The main remaining item is an overhaul of the data type system.

The meaning of each element within a NumPy array is described by its
datatype (dtype). NumPy arrays can hold all common numerical
datatypes, strings, datetimes, and generic Python objects, each of
which is identified by a dtype.  We are busy overhauling the datatype
system to make it behave consistently and to simplify the creation of
custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
While numerical dtypes make the foundation for most users,
there are a multitude of use cases which cannot prosper due to current
limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, and automatic
differentiation\cite{pyadolc}.
%Projects addressing these exist, each struggles with current
%limitations.  % TODO: I may need citations, we could link github issues.

@@

Now that we have made good progress on addressing technical debt, the project
is ready to grow its community of contributors and to better address the needs
of an ever-growing number of new users.  This is the main focus of the
forthcoming CZI grant.

>>>>

%NumPy is the fundamental array computation library for the Python
%ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Since its first stable release in October 2006, NumPy has unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
Not surprisingly, many important scientific research projects and
results crucially rely on NumPy.
It plays an essential role in many research analysis pipelines in
fields as diverse as physics, biology, astronomy, neuroscience,
material science, engineering, and chemistry.
% w/ citations
In particular, NumPy played a fundamental role in discovering
gravitational waves\cite{pycbc}, imaging black
holes\cite{eht-imaging}, and exploring the solar system (... Kepler? https://ui.adsabs.harvard.edu/abs/2012ascl.soft08004S/abstract).

%[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
%  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
%  IDL is now below Python, Pangeo <-> xarray, Earthcube,
%  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
%  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
%  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]

@@

%We showed that NumPy is actively developed according to a community
%vetted plan that integrates formal enhancement proposals.  Next, we
%give a more in-depth overview of the team developing NumPy, their
%activities, as well as the governance and community structures that
%guide the collaboration.

@@

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these 17 maintainers were active in 2019 and 4
of those were paid to work on the project full-time.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PRs) each.\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through github
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
With each release attracting more than a hundred new contributors.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
    \caption{Number of pull requests merged into the NumPy master branch for each
        quarter since 2012. The total number of PRs is indicated with the
        lower blue area showing the portion contributed by current or previous
        maintainers.}\label{fig:prs-over-time}
\end{figure}

The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
PRs merged into the NumPy master branch.
In this plot PRs opened by current and old maintainers are indicated.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.
Over the course of its history, NumPy has attracted PRs by 823 contributors.
However, its development relies heavily on a relatively small number
of active maintainers, who share more than half of the contributions among
themselves.

% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
% https://github.com/numpy/numpy/pull/6352
To formalize its unwritten development policy,
NumPy adopted an official Governance Document on October~5,
2015\cite{NumPyProjectGovernance}.
Project decisions are usually made by consensus of interested contributors.
This means for many thing everyone is entrusted with veto power.
A Steering Council, currently composed of 12~members, facilitates this
process and oversees daily development of the project by contributing code
and reviewing contributions from the community.

@@

The NumPy library consists of core functionality (e.g., the NumPy array data structure),
functions for manipulating arrays and doing scientific computation, and infrastructure
libraries.

@@

The \code{ndarray} data structure stores regularly spaced elements of a single
data type in a contiguous block of memory; along with meta-data such as shape,
it allows for the efficient representation of $n$-dimensional arrays.
More details about the data structure are given in ``The NumPy array:
a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.

The \emph{universal functions} (or \emph{ufuncs})
are functions written in C that implement efficient looping over
NumPy arrays. An important feature of ufuncs is the built-in
implementation of \emph{broadcasting}.  For example, the function
\code{arctan2(x, y)} is a ufunc that accepts two arguments and computes
$\tan^{-1}(y/x)$.  The ufunc loops over the dimensions of the inputs
in such a way that if, say, \code{x} is a 1-d array with length 3, and
\code{y} is a 2-d array with shape $2 \times 1$, the output will be
an array with shape $2 \times 3$.  The ufunc machinery takes care
of calling the function with all the appropriate combinations of
input array elements to compute the output array.
The elementary arithmetic operations of addition, multiplication, etc.,
are implemented as ufuncs, so broadcasting applies even in expressions
such as \code{x + y * z}.

@@

NumPy provides a large collection of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
It also provides extensive support for generating pseudorandom numbers
as well as an assortment of probability distributions.
For historical reasons, NumPy also includes basic functionality for
linear algebra, fast Fourier transforms and windowing,
and polynomial fitting.
We will continue supporting these, but we will not expand beyond them.

@@

NumPy also provides basic infrastructure for other packages in the scientific
Python ecosystem.
The \code{numpy.testing} subpackage provides functions such as
\code{assert\_allclose(actual, desired)} that may be used in unit
test suites for code that uses NumPy arrays.
The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
BLAS/LAPACK, and other relevant libraries for scientific computing.
The program \code{f2py} is a tool for
building NumPy-aware Python wrappers of Fortran functions.
NumPy itself does not use any Fortran code;  F2PY is part of NumPy
for historical reasons.

@@

The NumPy codebase has been actively developed for over two decades (first as
Numeric and NumArray).
Given the complexity of the codebase and the massive number of projects depending
on it, large changes require careful planning and substantial work.
Moreover, with the now prevalent need for distributed computing and the
emergence of specialized hardware such as GPUs, many new usecases have
arisen, which NumPy was not originally designed for.
With the recent infusion of funding and the newly formalized NEP process in
place we have been able to tackle a number of important large scale changes.
We highlight two of those below, as well as changes made to our testing
infrastructure to support hardware platforms used in large scale computing.

@@

A vast number of projects are built on NumPy;
these projects are consumers of the NumPy API.
%The central position of NumPy takes within the numerical and scientific
%Python ecosystem means that a vast number of projects are built on it.
%These projects are consumers of the NumPy API.
Over the last several years, a growing number of projects are providers of
a \emph{NumPy-like API} and array objects targeting audiences with specialized
needs beyond NumPy's capabilities.
For example, the NumPy API is implemented by several popular tensor computation
libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
and Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}}.
It is also implemented in packages that support sparse arrays
such as \code{scipy.sparse} and \code{pydata.sparse}.
Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.

The multitude of specialized projects creates the difficulty that consumers
of these NumPy-like APIs write code specific to a single project and do not support
all of the above array providers.
This is a burden for users relying on the specialized array-like, since
a tool they need may not work for them.
It also creates new challenges for end-users who need to transition
from NumPy to a more specialized array.
The growing multitude of specialized projects with NumPy-like APIs threatened
to again fracture the scientific Python community.

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.  For
example, here is how a CuPy GPU array can be passed through NumPy for
processing, with all operations being dispatched back to CuPy:

@@

\begin{lstlisting}
from numpy.random import Generator, PCG64, SeedSequence

seq = SeedSequence(10304245474441179933310169597522947680)
pcg = PCG64(seq)
gen = Generator(pcg)
\end{lstlisting}

@@

The final goal of the new API is to improve extensibility. \code{RandomState} is
a monolithic object that obscures all of the underlying state and functions. The
component architecture is one part of the extensibility improvements. The
underlying functions (written in C) which transform the output of a bit
generator to other distributions are available for use in CFFI. This allows the
same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
Numba. Both the bit generators and the low-level functions can also be used in C
or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
NumPy source. Alternative approaches that avoid this extra step are being
explored.}

@@

Most of the technical work for the Moore and Sloan grant is now complete.
The main remaining item is an overhaul of the data type system, which
we are currently working on \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
The meaning of each element within a NumPy array is described by its
datatype (dtype). NumPy arrays can hold all common numerical
datatypes, strings, datetimes, and generic Python objects, each of
which is identified by a dtype.
%We are busy overhauling the datatype
%system to make it behave consistently and to simplify the creation of
%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
While numerical dtypes make the foundation for most users,
there are a multitude of use cases which cannot prosper due to current
limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, and automatic
differentiation\cite{pyadolc}.
%Projects addressing these exist, each struggles with current
%limitations.  % TODO: I may need citations, we could link github issues.
The proposed overhaul of the datatype system will make it behave consistently and
to simplify the creation of custom dtypes, both in C and in Python

@@

Having made progress on addressing technical debt accrued over two decades of
development and modernizing the code to handle new use cases, the project
is ready to grow its community of contributors and to better address the needs
of an ever-growing number of new users.  This is the main focus of the
forthcoming CZI grant.

### 2019-12-20 paper.tex
Proofread manuscript

\lstset{ %
  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
  breaklines=true,   % automatic line breaking only at whitespace
  captionpos=b,   % sets the caption-position to bottom
  commentstyle=\color{gray},  % comment style
  keywordstyle=\color{blue},  % keyword style
  stringstyle=\color{red},  % string literal style
  upquote=true  %straight single quotes (requires textcomp)
}

@@

\begin{abstract}
  NumPy is the fundamental array computing library for Python.
  It has been actively developed for over two decades and has
  more than 850 contributors, nearly 300,000 dependent repositories
  on GitHub, and millions of downloads per year.
  It is extensively used in research and teaching.
  According to a 2018 survey, NumPy is the second most commonly used machine
  learning tool among enterprises \cite{451report2018}.
  NumPy continues to evolve to meet the needs of its community, to
  remain relevant in an ever-changing computing landscape, and to be a
  fundamental pillar of support for the next decade of data science in
  Python.
  In this article, we provide an overview of the capabilities and development
  practices of the library and highlight some recent technical
  developments.
\end{abstract}

>>>>

\lstset{ %
  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
  xleftmargin=2\parindent,
  breaklines=true,   % automatic line breaking only at whitespace
  captionpos=b,   % sets the caption-position to bottom
  commentstyle=\color{gray},  % comment style
  keywordstyle=\color{blue},  % keyword style
  stringstyle=\color{red},  % string literal style
  upquote=true  %straight single quotes (requires textcomp)
}

@@

\begin{abstract}

  NumPy is the fundamental array computation library for the Python ecosystem
  \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
  It has been actively developed for over two decades and has
  more than 850 contributors, nearly 300,000 dependent repositories
  on GitHub, and millions of downloads per year.
  It is extensively used in research and teaching.
  According to a 2018 survey, NumPy is the second most commonly used machine
  learning tool among enterprises \cite{451report2018}.
  NumPy continues to evolve to meet the needs of its community, to
  remain relevant in an ever-changing computing landscape, and to be a
  fundamental pillar of support for the next decade of data science in
  Python.
  In this article, we provide an overview of the project and library
  organization,  highlight some recent technical developments, and
  outline are future plans.
\end{abstract}
